Refactoring `nls()`: a constructive critique

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
convert paper to rjtools

>>>>

# The `nls()` function: strengths and shortcomings

`nls()` is the tool in base R, the primary software software distribution
from the Comprehensive R Archive Network (*https://cran.r-project.org*),
for estimating nonlinear statistical models. The function dates to the 1980s and the work
related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).

The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
nonlinear models that are expressed as **formulas**. In particular, we note that it

- handles formulas that include R functions, even ones which call calculations in
  other programming languages
- allows data to be weighted or subset
- can estimate bound constrained parameters
- provides a mechanism for handling partially linear models
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded (`selfStart`) models

With such a range of features and long history, the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
as we suggest below.

## Feature: Convergence and termination tests (FIXED)

A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
(i.e., perfect fit) data was corrected thanks to suggestions by one of us.

In the manual page for `nls()` in R 4.0.0 there is the warning:

>  **Do not use `nls` on artificial "zero-residual" data.**

>  The `nls` function uses a relative-offset convergence criterion
  that compares the numerical imprecision at the current parameter
  estimates to the residual sum-of-squares.  This performs well on data of
  the form

>  $$  y = f(x, \theta) + eps $$

>  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form

>  $$  y = f(x, \theta)  $$

>  because the criterion amounts to comparing two components of the round-off 
  error.

>  If you wish to test `nls` on artificial data please add a noise component, 
  as shown in the example below.

This amounted to admitting R cannot solve well-posed problems unless data is polluted
with errors.

This issue can be easily resolved. The "termination test" for the **program** rather than for 
"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
and estimates how much the sum of squares loss function should decrease. This estimate is divided 
by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

Adding a small quantity to the loss function before dividing avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
with a default value of zero. Setting it to a small number -- 1.0 is 
a reasonable choice --
allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.

# References
